---
title: "Perform m/z and RT(I) matching with masstrixR"
author: "Dr. Michael Witting"
date: "23 January 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Retention times are useful orthogonal information to m/z values for metabolite identification. <code>masstrixR</code> can combine m/z and RT search, when a suitable database is supplied. In this case the SQLite database that is used has to contain retention time information. The example below shows how a suitable compound list for combined m/z and RT search looks like and how it is used with <code>masstrixR</code>. The example data is taken from Stoffel et al. [1] and contains a small database of substances and their retention time indicies (RTIs) obtained on a reversed phase separation. The databsae is based on authentic chemical standards and unknown samples are derived from measurements of C. elegans metabolite extracts.

```{r read compound list}
#load required library
library(readr)
library(masstrixR)

# get example file from package
exampleFile <- system.file("extdata", "exampleData\\Celegans_mz_rt\\rtidb.txt", package = 'masstrixR')

# read file into tibble (the funciton from readr is used)
compoundList <- read_tsv(exampleFile, col_types = cols(exactmass = col_double(),
                                                       rt = col_double()))

head(compoundList)
```

In order to work with retention times the argument <code>rt</code>, which is by default set to <code>FALSE</code>, has to be set to <code>TRUE</code>. This will check if all entries in the compound list contain a RT value. This is necessary, because the database search also searches for a matching RT.

```{r generate sqlite}
# adducts used for DB generation
adducts <- c("[M+H]+", "[M+Na]+")

# create compound list for DB creation
newCompoundList <- prepareCompoundList(compoundList, adductList = adducts, rt = TRUE)

# check if compound list is valid and create SQLite DB
if(validateCompoundList(newCompoundList)) {
  dbFileName <- createDb(newCompoundList, "example_pos_MH_MNa_withRT")
}
print(dbFileName)
```

Next, the example data is loaded. It contains one column called <code>RT</code>, which contains the caluclated RTIs for the samples and one column <code>RT.min</code> with the original retention time in minutes. The database search is performed as simultaneous search for fitting m/z-RT pairs. The allowed retention time deviation is defined in the <code>rtTol</code> field, while the type is either defined as <code>"abs"</code> or <code>"rel"</code> for absolute or relative error respectively. In the used example an absolute RTI error of 5 units is used. Please not that no testing if the RTs in database and in the data are in the same unit is performed.

```{r}
peakTableFile <- system.file("extdata", "exampleData\\Celegans_mz_rt\\peakTable.txt", package = 'masstrixR')
peakTable <- read_tsv(peakTableFile)

#annotate
annotationResults <- mzSearch(peakTable, dbFileName,
                              mzTol = 0.005, mzTolType = "abs",
                              rt = TRUE, rtTol = 5, rtTolType = "abs")

head(annotationResults)
```


## Literature

[1] Stoffel et al. 2019
