---
title: "Create SQLiteDBs for masstrixR"
author: "Michael"
date: "30 November 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

<code>masstrixR</code> performs annotation of MS^1^ data with putative metabolites. In order to perfrom this annotation a fitting database of metabolites is required. <code>masstrixR</code> uses the SQLite database framework for fast and efficient searching. Measured masses are compared with theoretical adduct masses of metabolites of interest withing a certain mass error, either defined as absolute error in Da or relative error in ppm.
Databases for <code>masstrixR</code> are generated from compound lists and adduct definitions. <code>masstrixR</code> contains calculation rules for several possible adducts. Adducts are referenced by their normal nomenclature, e.g. [M+H]+.

## Working with adducts

The adduct calculation rules are defined in a list and contain an additive and multiplicative part, as well as subtractive and additive part for the generation of ion sum formulas. The function <code>getAdductCalc()</code> returns a list with this calculation rules.

```{r adducts}
library(masstrixR)
adductCalc <- getAdductCalc()
adductCalc["[M+H]+"]
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
