---
title: "Using masstrixR - Perform MS1 Search"
author: "Michael"
date: "19 Oktober 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Annotation of LC-MS data

The database can be now used for annotation of MS^1^ data from different experiments. The supplied example file reads a .gda file exported from Genedata Expressionist for MS 11.0. A list with column and row annotations as well as the actual data is returned.

```{r read .gda file}
# read example .gda file
exampleGDA <- readGdaFile("..\\example\\NaAcHILICPos_Cluster.gda")

# get row annotations with m/z values
rowAnno <- exampleGDA[[3]]
rowAnno$ClusterName <- row.names(rowAnno)

#annotate
annotationResults <- mzSearch(rowAnno, dbFileName, mzTol = 0.005, mzTolType = "abs")

```

## Fast coding

If no persistent .sqlite file is needed or only a fast lookup of specific m/z values shall be performed the <code>mzLookUp()</code> function can be used to perform the same code as above in a single line.

```{r}
# perform m/z look up
annotationResultsLookup <- mzLookUp(rowAnno, newCompoundList, mzTol = 0.005, mzTolType = "abs")
```
