---
title: "Using masstrixR - Create a SQL database"
author: "Michael"
date: "3 August 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# MassTRIX in R - Annotating MS1 data made easy...

## Introduction

Annotation of MS^1^ data with potential metabolites from different databases. This achieved by calculating theoretical adduct masses from metabolites and compare measured m/z values with theoretical ones within a give error (in Da or ppm).
The masstrixR package uses SQLite for back-end databases storing theoretical adduct masses of a given metabolite database (e.g. YMDB). Two possibilities exist. Either a persistent database stored in a single .sqlite file can be generated and re-used or a in-memory database can be generated "on-the-fly". Additionally, large database stored in a .sqlite file can be read into memory for enhanced performance.

## Reading a DB

Metabolite databases from which adduct databases shall be created have to be supplied as tab separated text file with the following headers:

id	smiles	inchi	inchikey	formula	name	exactmass

Header names have to be exactly the same as stated above.

```{r read compound list}
library(masstrixR)

# read YMDB text file and create example .sqlite fil
compoundList <- data.frame(read.table("..\\DBs\\Txt\\chebi_20181015.txt", sep = "\t", header = T, stringsAsFactors = F, comment.char = "~"))

#head(compoundList)
```

## Generate SQLite database and store to file

A persistent .sqlite database file containing pre-calculated adduct masses can be generated from this compound list using functions from masstrixR. The adducts that shall be covered in this database have to be defined before. The <code>getAdductNames()</code> function returns a vector with all adduct names used in <code>masstrixR</code>.
The following example creates a DB with [M+H]^+^ and [M+Na]^+^ adducts. The <code>createDb()</code> function creates a .sqlite file. A data path to which the SQLite shall be safed with a name has to be supplied. After succesfull creation the data path is returned.


```{r generate sqlite}
# list all adducts
getAdductNames()

# adducts used for DB generation
adducts <- c("M+H", "M+Na")

# create compound list for DB creation
newCompoundList <- prepareCompoundList(compoundList, adductList = adducts)

# create SQLite DB
dbFileName <- createDb(newCompoundList, "..\\DBs\\SQLite\\chebi_20181015_pos_MH_MNa")
```
