---
title: "Perform m/z matching with masstrixR"
author: "Dr. Michael Witting"
date: "23 January 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introdcution

MassTRIX was build for the annotation of m/z values from high resolution mass spectrometry with putative metabolites. Different versions of MassTRIX have been published as online server [1-3]. The new version of MassTRIX is implemented as R package, called <code>masstrixR</code>. The main purpose is to provide annotation to m/z features and allow analysis of MS^2^ for metabolite identification.

This vignette describes the basic workflow for annotation of m/z values obtained from MS^1^. The input can be derived from direct infusion or chromatographic experiments. <code>masstrixR</code> uses SQLite database files to store metabolite databases and calculated adducts. The generation of such a database is explained in the vignette "Create SQLiteDBs for masstrixR". 

## Loading a database

For the following example we load an example database file installed with <code>masstrixR</code>. This database contains precalculated adducts ([M+H]+ and [M+Na]+) for a selection of metabolites. The database file is simply loaded via its file name and only opened on demand.

```{r load DB}
dbFileName <- system.file("extdata", "exampleData\\databases\\example_pos_MH_MNa.sqlite", package = 'masstrixR')
```

## Reading m/z data

Data that shall be annotated has to be supplied as data frame with column headers. The column containing the m/z values is identified by its name. Currently supported header names are:

* mz (<code>$mz</code>)
* m.z (<code>$m.z</code>)
* mzmed (<code>$mzmed</code>)

In this example a .gda file from Genedata Expressionist for MS is loaded using the <code>readGdaFile</code> function, but any other data frame contain the m/z values will work. In future more distinct reader functions will be implemented, e.g. for Agilent .cef files etc. The function <code>readGdaFile</code> returns a list with the acutal intensity values, the annotation of the samples and the annotation of the features, which we require for m/z annotation. This data frame is the third in the list.

```{r read .gda file, message=FALSE}
# load masstrixR
library(masstrixR)

# read example .gda file
gdaFile <- system.file("extdata", "exampleData\\Celegans_mz\\NaAcHILICPos_Cluster.gda", package = 'masstrixR')
exampleGDA <- readGdaFile(gdaFile)

# get row annotations with m/z values
rowAnno <- exampleGDA[[3]]
rowAnno$ClusterName <- row.names(rowAnno)
head(rowAnno)
```

## Perform m/z annotation

After all the data is available, the m/z annotation can be performed. This is done with the function <code>mzSearch</code>. It requires some arguments. First, the data frame for which the annotation shall be performed (in this case <code>rowAnno</code>), the path to the .sqlite file containing the database (<code>dbFileName</code>) and the tolerance for the mass search. The tolerance can be given either as absolute value (<code>"abs"</code>) in Da or as relative error (<code>"ppm"</code>) in ppm. The argument <code>mzTol</code> defines the maximum tolerance. The function <code>mzSearch</code> returns a data frame with all features that have been annotated. It contains all the original columns plus all columns from the database.

```{r}
#annotate
annotationResults <- mzSearch(rowAnno, dbFileName, mzTol = 0.005, mzTolType = "abs")

head(annotationResults)
```

If large database shall be used for annotation the database file can be first read into the memory for improved performance. This is done by adding the argument <code>mode</code>. The default value for this argument is <code>"onDisk"</code>. The value <code>"inMemory"</code> creates a copy of the database in memory, which is used for annotation.

```{r}
#annotate with DB in memory
annotationResults <- mzSearch(rowAnno, dbFileName, mode = "inMemory", mzTol = 0.005, mzTolType = "abs")
```

The function <code>mzSearch</code> contains several other options for more advanced workflows, which include RT and CCS matching. They are explained in the respective vignettes.

## Literature

[1] Suhre & Schmitt-Kopplin
[2] Wägele et al.
[3] Witting & Schmitt-Kopplin
