---
title: "Use of MS2 functions"
author: "Michael"
date: "7 Februar 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Analysis of MS^2^ data can involve several steps beyond matching with a database. Different approaches can be used to analyze tandem MS spectra. One of the most common is to search for known neutral losses from the precursor masses, specific fragment m/z or mass differences between fragments. <code>masstrixR</code> offers functions to execute all of this steps. The example data contains spectra of ceramides from Hänel et al. (2019, under review) detected in C. elegans. These ceramides contain a unusual C17iso sphingoid base.

First we read the spectra from the .mgf file using functions from the <code>MSnbase</code> package. <code>masstrixR</code> uses <code>Spectrum2</code> for working with MS^2^ spectra. Individual spectra are best stored in a <code>Spectra</code> object which allows to add arbitary annotations using the <code>mcols()</code> function.

```{r read ms2 spectra}
# load required library
library(masstrixR)
library(MSnbase)

# get example file from package
mgfFile <- system.file("extdata", "exampleData\\Celegans_lipids_ms2\\ceramides.mgf", package = 'masstrixR')

# read mgf file with MS2 spectra and isolate all MS2 spectra a list
ms2data <- readMgfData(mgfFile)
ms2spectra <- Spectra(spectra(filterMsLevel(ms2data, msLevel = 2)))

# working with logical statements for fragments
mcols(ms2spectra)$waterLoss <- unlist(lapply(ms2spectra, containsNeutralLossIon, neutralLossMass = 18.010565))
mcols(ms2spectra)$c17sphingoid <- unlist(lapply(ms2spectra, containsProductIon, productIonMz = c(250.2529, 268.2635, 238.2530), multiplePi = "all"))
mcols(ms2spectra)$c27ohFa <- unlist(lapply(ms2spectra, containsFragmentDifference, fragmentMassDifference = 392.401816))

# check the invidual columns
mcols(ms2spectra)$c17sphingoid
mcols(ms2spectra)$c27ohFa

# make combined searches
selectedSpectrum <- ms2spectra[which(mcols(ms2spectra)$waterLoss & 
                   mcols(ms2spectra)$c17sphingoid &
                   mcols(ms2spectra)$c27ohFa)]

# plot
plotSpectrum(selectedSpectrum[[1]], plotIt = TRUE, highlight = TRUE, highlightMz = c(250.2529, 268.2635, 238.2530))
```

```{r}
# get example file from package
ms2db <- system.file("extdata", "exampleData\\Celegans_lipids_ms2\\wormLipidDb_MS2.sqlite", package = 'masstrixR')

# adducts to search for
adducts <- c("[M+H]+", "[M+Na]+")

# empty data frame for results
result <- data.frame()

for(i in seq_along(ms2spectra)) {
  
    # create empty Spectra object for results
  searchResult <- new("Spectra")
  
  # perform precursor search
  # iterate over adducts
  for(adduct in adducts) {
    searchResultClipboard <-
      masstrixR::searchByPrecursor(
        precursorMz(ms2spectra[[i]]),
        ms2db,
        precursorType = adduct,
        mzTol = 0.005
      )
    
    # if 1 or more results were found, add to searchResult
    if(length(searchResultClipboard) > 0) {
      searchResult <- append(searchResult, searchResultClipboard)
    }
  }
  
  result <- rbind.data.frame(result, createResultsSet(ms2spectra[[i]], searchResult, prefix = "test", storePlot = FALSE))
}
```

