---
title: "Using masstrixR - Create a SQLite database and perform MS1 search"
author: "Michael"
date: "3 August 2018"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using masstrixR - Create a SQLite database}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# MassTRIX in R - Annotating MS1 data made easy...

## Introduction

Annotation of MS^1^ data with potential metabolites from different databases. This achieved by calculating theoretical adduct masses from metabolites and compare measured m/z values with theoretical ones within a give error (in Da or ppm).
The masstrixR package uses SQLite for back-end databases storing theoretical adduct masses of a given metabolite database (e.g. YMDB). Two possibilities exist. Either a persistent database stored in a single .sqlite file can be generated and re-used or a in-memory database can be generated "on-the-fly". Additionally, large database stored in a .sqlite file can be read into memory for enhanced performance.

## Reading a DB

Metabolite databases from which adduct databases shall be created have to be supplied as tab separated text file with the following headers:

id	smiles	inchi	inchikey	formula	name	exactmass

Header names have to be exactly the same as stated above.

```{r read compound list}
library(masstrixR)

chebiExample <- system.file("extdata", "chebi_20181015.txt", package = 'masstrixR')

# read YMDB text file and create example .sqlite fil
compoundList <- data.frame(read.table(chebiExample, sep = "\t", header = T, stringsAsFactors = F, comment.char = "~"))
```

## Generate SQLite database and store to file

A persistent .sqlite database file containing pre-calculated adduct masses can be generated from this compound list using functions from masstrixR. The adducts that shall be covered in this database have to be defined before. The <code>getAdductNames()</code> function returns a vector with all adduct names used in <code>masstrixR</code>.
The following example creates a DB with [M+H]^+^ and [M+Na]^+^ adducts. The <code>createDb()</code> function creates a .sqlite file. A data path to which the SQLite shall be safed with a name has to be supplied. After succesfull creation the data path is returned.


```{r generate sqlite}
# list all adducts
getAdductNames()

# adducts used for DB generation
adducts <- c("M+H", "M+Na")

# create compound list for DB creation
newCompoundList <- prepareCompoundList(compoundList, adductList = adducts, rt = FALSE, ccs = FALSE, extId = FALSE)

# check if compound list is valid and create SQLite DB
if(validateCompoundList(newCompoundList, rt = FALSE, ccs = FALSE)) {
  dbFileName <- createDb(newCompoundList, "chebi_20181015_pos_MH_MNa")
}
print(dbFileName)
```

## Annotation of LC-MS data

The database can be now used for annotation of MS^1^ data from different experiments. The supplied example file reads a .gda file exported from Genedata Expressionist for MS 11.0. A list with column and row annotations as well as the actual data is returned.

```{r read .gda file, message=FALSE}
# read example .gda file
gdaFile <- system.file("extdata", "NaAcHILICPos_Cluster.gda", package = 'masstrixR')
exampleGDA <- readGdaFile(gdaFile)

# get row annotations with m/z values
rowAnno <- exampleGDA[[3]]
rowAnno$ClusterName <- row.names(rowAnno)

#annotate
annotationResults <- mzSearch(rowAnno, dbFileName, mzTol = 0.005, mzTolType = "abs")

```

## Fast coding

If no persistent .sqlite file is needed or only a fast lookup of specific m/z values shall be performed the <code>mzLookUp()</code> function can be used to perform the same code as above in a single line.

```{r, message=FALSE}
# perform m/z look up
annotationResultsLookup <- mzLookUp(rowAnno, newCompoundList, mzTol = 0.005, mzTolType = "abs")
```
